---
title: "steveHMM_vignette"
author : Sidonie Foulon, Thérèse Truong, Anne-Louise Leutenegger, Hervé Perdry
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{steveHMM_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

**steveHMM** package contain several functions. The aim is to estimate the Hidden Markov model (HMM) parameters via different methods. The four main functions corresponds to four maximum likelihood algorithms : \

* **quasi_newton** : direct maximisation of the likelihood with a quasi-Newton algorithm (L-BFGS-B)\
* **EM** : Baum-Welch algorithm, an Expectation-Maximisation (EM) algorithm \
* **SQUAREM** : Squared iterative method, an accelerated EM algorithm (Varadhan, 2008) \
* **QNEM** : our novel hybrid algorithm, combines the Baum-Welch and quasi-Newton BFGS algorithm. \

We will show how to use each of these four functions on an example.

# Example

We will use a simple example based on a fictitious situation. Every day of the year, in a secret underground installation, a security guard observes whether the director comes in with an umbrella or not (Gao, 2021). During 56 days, the security guard keep track of his observations on his note pad as follows: U if the director carries an umbrella; N otherwise.  
The guards aims to predict the daily weather conditions (hidden states) based on the umbrella observations: rainy (noted R) or dry (noted D) day. Here the hidden state is the weather.  
Since the security guard is in a underground installation, he can not observe directly the weather: he has to rely only on his observations of the umbrella status to determine the weather. The parameters to estimate are *a* the probability of weather state transition and *b* the probability of error in the umbrella status (not carrying an umbrella on rainy day or carrying an umbrella on dry day). The transition and emission matrix are presented respectively in the following Tables.

| Transition matrix | $S_i = D$ | $S_i = R$ |
|:-----------------:|:---------:|:---------:|
|   $S_{i-1} = D$   |    1-a    |     a     |
|   $S_{i-1} = R$   |     a     |    1-a    |


| Emission matrix | $S_i = D$ | $S_i = R$ |
|:---------------:|:---------:|:---------:|
|    $X_i = N$    |    1-b    |     b     |
|    $X_i = U$    |     b     |    1-b    |

Here are the observations *X* for our example (1 code for "No umbrella" and 2 for "Umbrella") and the initial parameters *par*.

```{r}
X <- c(1, 1, 2, 2, 2, 2, 2, 2, 2, 1, 2, 2, 2, 1, 2, 2, 1, 1, 1, 1,
    1, 1, 1, 1, 1, 2, 1, 1, 1, 1, 1, 1, 1, 2, 1, 1, 1, 2, 2, 2, 1,
    2, 2, 1, 2, 2, 1, 2, 2, 2, 2, 2, 1, 2, 2, 2)

par <- c(a = 0.3, b = 0.15)

```


# Model

Before running our functions, we have to create the model of the example. The function **model** takes the set of parameters *theta*, the observations *obs* and the names of the hidden states *name.S*. It returns in a list, the transition matrix *trans*, the stationary probability *pi* and a matrix of the emission probabilities for each observation *p.emiss*.

*trans* is constructed from the transition parameter *a*. 
In this example, the stationary state does not depend on any of the two parameters. For some examples, the stationary state may be computed from the parameters.
*p.emiss* is constructed from the emission parameter *b*. It is a matrix of the same number of columns as the length of *obs* (here 56) and one rows by hidden state (here 2).

```{r}
model <- function(theta, obs, name.S = c("Dry", "Rainy")) {
  
  # parameter of transition matrix 
  a <- theta[1]
  trans <- matrix( c(1-a, a, a, 1-a), byrow = TRUE, nrow = 2, ncol = 2)
  colnames(trans) <- rownames(trans) <- name.S
  
  # stationary state
  stat <- c(0.5, 0.5)
  
  # parameter of emission matrix
  b <- theta[2]
  p.emiss <- rbind(ifelse(obs == 1, 1-b, b), ifelse(obs == 1, b, 1-b))
  rownames(p.emiss) <- name.S
  
  #return
  list(trans = trans, pi = stat, p.emiss = p.emiss)
}

mod <- model(theta = par, obs = X)
mod

```




This **model** function will be used in all four main functions.


# M-step

The M-step corresponds to the maximisation step of the EM algorithm. This step depends on the model so can not be generalised, on the contrary of the E-step. The **M.step** function takes the observations *obs* and the backward results *backward* from the forward-backward algorithm used in E-step. The backward can be computed with the **backward** function of the **steveHMM** package. This function takes the results of the **forward** algorithm, obtained with the so-called function.

```{r setup, message = FALSE}
library(steveHMM)
```

```{r}
fw <- forward(modele = mod)

bw <- backward(modele = fw)

```

Here is how is constructed a M-step function. The parameter *a* corresponds to the probability of change in the hidden state, ie. switching from a dry day to a rainy day and vice versa. In the results of the backward algorithm, $\delta_i(s,t) = \mathbb{P}_{\theta}(S_{i-1}=s, S_i=t | X_1^{L} = x_1^{L})$. The probability of change will be the sum of prbablities $\delta_i(D,R)$ and $\delta_i(R,D)$ on the total number of transitions.
For the parameter *b*, the parameter of error in the umbrella status, ie. bringing an umbrella on a dry day or forget it on a rainy day, we will use another backward algorithm result. We rely here on the probability $\varphi_i(s) = \mathbb{P}_{\theta}(S_i=s | X_1^{L} = x_1^{L})$. To estimate *b*, we sum the probabilities $\varphi_i(R)$ only on the observations coded 1 for "No umbrella", and $\varphi_i(D)$ only on the observations coded 2 for "Umbrella", divided by the total number of observations.

The **M.step** function returns the re-estimated parameters.

```{r}
M.step <- function(obs, backward) {
  l <- ncol(backward$phi)
  
  # parameter a
  a <- (sum(backward$delta[1,2,] + backward$delta[2,1,]))/(l-1)
  
  # parameter b
  b <- (sum(backward$phi[2,which(obs == 1)]) + sum(backward$phi[1,which(obs == 2)]))/l
  
  c(a,b)
}

Mstep <- M.step(obs =X, backward = bw)
Mstep

```

This function will be used for all four main functions, except the **quasi_newton** function.

# quasi_newton function

The **quasi_newton** function let us estimate the HMM parameters with direct maximisation of the likelihood. We use a quasi-Newton type method called L-BFGS-B. In this function we are using the R function **optim**. 
**quasi_newton** function takes the initial parameters *theta*, the observations *obs*, the model function previously constructed in *modele.fun*. If you want to keep track of the different estimations of the parameters, you can use the argument *trace*. You can also add the arguments specific to the **optim** function. Here we added the arguments *lower* and *upper* related to the bounds for the estimations of the parameters *a* and *b*.

```{r}
qn <- quasi_newton(theta = par, obs = X, modele.fun = model, trace = TRUE, lower = c(0.01,0.01), upper = c(0.99,0.99))
qn
```

This function returns the same output as the **optim** function, more the number of forward algorithm calls.

# EM function

With this function, we can apply the Baum-Welch algorithm, a variant of the EM algorithm. **EM** function takes the initial parameters *theta*, the observations *obs*, the model and M.step functions previously constructed in respectively *modele.fun* and *M.step.fun*. There also are optional arguments such as *max.iter* to bound the number of iterations of the algorithm or *trace.theta* to keep track of the different estimations of the parameters. By default, the stopping criterion is *reltol*, but it can be switched to a threshold stopping criterion with "*criteria* = 'epsilon' ".

```{r}
em <- EM(theta = par, obs = X, modele.fun = model, M.step.fun = M.step, max.iter = 100, trace.theta = TRUE, criteria = "reltol")
em
```

The function returns a list with the final estimation of the parameters *theta*, the number of iterations *iter*, the *likelihood* and if the *trace.theta* option is on, the track of all parameters estimated in *Theta*.

# SQUAREM function

The SQUAREM function is used to apply the accelerated EM algorithm SQUAREM. Its arguments are the same as **EM** function. You can also bound the space of parameters with *lower* and *upper* arguments.

```{r}
squarem <- SQUAREM(theta = par, obs = X, modele.fun = model, M.step.fun = M.step, lower = c(0,0), upper = c(1,1), max.iter = 100, trace.theta = TRUE, criteria = "reltol")
squarem
```

Additional to the values returned by the **EM** function, **SQUAREM** returns the number of *forward* and *backward* algorithms performed.

# QNEM function

The last function, we are presenting here is the **QNEM** function, an implementation of our algorithm which mix the Baum-Welch EM and the quasi-Newton BFGS algorithms. The mandatory arguments are the same as the **EM** function. Yet, you can also add some optional arguments, such as those used by the **SQUAREM** function, but also you can set the number of Baum-Welch algorithm to perform before switching to BFGS algorithm (here, we chose 3, ie. the algorithm will perform 3 iterations of the Baum-Welch algorithm, then only the BFGS algorithm, unless the curvature condition is not met anymore). To display information while running the algorithm, you can add "*verbose* = TRUE".

```{r}
qnem <- QNEM(theta = par, obs = X, modele.fun = model, M.step.fun = M.step, max.iter = 100, upper = c(1,1), lower = c(0,0), trace.theta = TRUE, nb.em = 3, verbose = TRUE)
qnem
```

The function returns the same output as the **SQUAREM** function.


